---
title: "tidyverse and data.table"
output: pdf_document
---

```{r}
library(pacman)
p_load(nycflights13, tidyverse)
```

Tidyverse: dplyr and tidyr

# dplyr

# 1. filter

Filter (i.e. subset) rows based on their values.

$~$

```{r}
starwars %>% 
  filter( 
    species == "Human", 
    height >= 190
    )
```

$~$

```{r}
starwars %>% 
  filter(grepl("Skywalker", name))
```

$~$

```{r}
starwars %>% 
  filter(is.na(height))
```

$~$

```{r}
starwars %>% 
  filter(!is.na(height))
```

# 2. arrange

Arrange (i.e. reorder) rows based on their values.

$~$

```{r}
starwars %>% 
  arrange(birth_year)
```

$~$

```{r}
starwars %>% 
  arrange(desc(birth_year))
```


# 3. select

Select (i.e. subset) columns by their names:

$~$

```{r}
starwars %>% 
  select(name:skin_color, species, -height)
```

$~$

```{r}
starwars %>%
  select(alias=name, crib=homeworld, sex=gender)
```

$~$

```{r}
starwars %>% 
  select(name, contains("color"))
```

$~$

The select(..., everything()) option is another useful shortcut if you only want to bring some variable(s) to the "front" of a data frame.

```{r}
starwars %>% 
  select(species, homeworld, everything()) %>%
  head(5)
```

# 4. mutate

Create new columns.

$~$

```{r}
starwars %>% 
  select(name, birth_year) %>%
  mutate(dog_years = birth_year * 7) %>%
  mutate(comment = paste0(name, " is ", dog_years, " in dog years."))
```

$~$

same as the last one. We can chain multiple mutates in a single call.

```{r}
starwars %>% 
  select(name, birth_year) %>%
  mutate(
    dog_years = birth_year * 7, ## Separate with a comma
    comment = paste0(name, " is ", dog_years, " in dog years.")
    )
```

$~$

```{r}
starwars %>% 
  select(name, height) %>%
  filter(name %in% c("Luke Skywalker", "Anakin Skywalker")) %>% 
  mutate(tall1 = height > 180) %>%
  mutate(tall2 = ifelse(height > 180, "Tall", "Short"))
```

$~$

Lastly, combining mutate with the new across feature in dplyr 1.0.0+ allows you to easily work on a subset of variables. For example:

```{r}
starwars %>% 
  select(name:eye_color) %>% 
  mutate(across(where(is.character), toupper)) %>%
  head(5)
```
# 5. summarise

Collapse multiple rows into a single summary value.

```{r}
starwars %>% 
  group_by(species, gender) %>% 
  summarise(mean_height = mean(height, na.rm = TRUE))
```

$~$

Note that including "na.rm = TRUE" (or, its alias "na.rm = T") is usually a good idea with summarise functions. Otherwise, any missing value will propogate to the summarised value too.

```{r}
starwars %>% 
  summarise(mean_height = mean(height))
```

$~$

```{r}
starwars %>% 
  summarise(mean_height = mean(height, na.rm = TRUE))
```

$~$

```{r}
starwars %>% 
  group_by(species) %>% 
  summarise(across(where(is.numeric), mean, na.rm=T)) %>%
  head(5)
```
# join

```{r}
left_join(flights, planes) %>%
  select(year, month, day, dep_time, arr_time, carrier, flight, tailnum, type, model)
```

# tidyr

# pivot_longer

Pivot wide data into long format (i.e. "melt").

```{r}
stocks = data.frame( ## Could use "tibble" instead of "data.frame" if you prefer
  time = as.Date('2009-01-01') + 0:1,
  X = rnorm(2, 0, 1),
  Y = rnorm(2, 0, 2),
  Z = rnorm(2, 0, 4))
stocks
```

$~$
```{r}
stocks %>% pivot_longer(-time, names_to="stock", values_to="price")
```

$~$

```{r}
tidy_stocks = 
  stocks %>% 
  pivot_longer(-time, names_to="stock", values_to="price")
tidy_stocks
```




# pivot_wider

Pivot long data into wide format (i.e. "cast").2

```{r}
tidy_stocks %>% pivot_wider(names_from=stock, values_from=price)

```

$~$

```{r}
tidy_stocks %>% pivot_wider(names_from=time, values_from=price)
```

# separate

Separate (i.e. split) one column into multiple columns.

```{r}
economists = data.frame(name = c("Adam.Smith", "Paul.Samuelson", "Milton.Friedman"))
economists
```

$~$

```{r}
economists %>% separate(name, c("first_name", "last_name"))
```


$~$

```{r}
jobs = data.frame(
  name = c("Jack", "Jill"),
  occupation = c("Homemaker", "Philosopher, Philanthropist, Troublemaker") 
  ) 
jobs
```

$~$

```{r}
jobs %>% separate_rows(occupation)
```

# unite 

Unite (i.e. combine) multiple columns into one.

```{r}
gdp = data.frame(
  yr = rep(2016, times = 4),
  mnth = rep(1, times = 4),
  dy = 1:4,
  gdp = rnorm(4, mean = 100, sd = 2)
  )
gdp
```
$~$

```{r}
gdp %>% unite(date, c("yr", "mnth", "dy"), sep = "-")
```
$~$

```{r}
gdp_u = gdp %>% unite(date, c("yr", "mnth", "dy"), sep = "-") %>% as_tibble()
gdp_u
```






























