---
title: "webscraping_CSS"
output: pdf_document
---

#  (1) Server-side and CSS




```{r}
## Install development version of rvest if necessary
if (numeric_version(packageVersion("rvest")) < numeric_version('0.99.0')) {
  remotes::install_github('tidyverse/rvest')
}
## Load and install the packages that we'll be using today
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, rvest, lubridate, janitor, data.table, hrbrthemes)
## My preferred ggplot2 plotting theme (optional)
theme_set(hrbrthemes::theme_ipsum())
```

```{r}
hrbrthemes::import_roboto_condensed()
```


# Webscraping basics

The next two lectures are about getting data, or “content”, off the web and onto our computers. We’re all used to seeing this content in our browers (Chrome, Firefox, etc.). So we know that it must exist somewhere. However, it’s important to realise that there are actually two ways that web content gets rendered in a browser:

1. Server-side
2. Client side
You can read here for more details (including example scripts), but for our purposes the essential features are as follows:

# 1. Server-side

The scripts that “build” the website are not run on our computer, but rather on a host server that sends down all of the HTML code.
E.g. Wikipedia tables are already populated with all of the information — numbers, dates, etc. — that we see in our browser.
In other words, the information that we see in our browser has already been processed by the host server.
You can think of this information being embeded directly in the webpage’s HTML.
Webscraping challenges: Finding the correct CSS (or Xpath) “selectors”. Iterating through dynamic webpages (e.g. “Next page” and “Show More” tabs).
Key concepts: CSS, Xpath, HTML


# 2. Client-side

The website contains an empty template of HTML and CSS.
E.g. It might contain a “skeleton” table without any values.
However, when we actually visit the page URL, our browser sends a request to the host server.
If everything is okay (e.g. our request is valid), then the server sends a response script, which our browser executes and uses to populate the HTML template with the specific information that we want.
Webscraping challenges: Finding the “API endpoints” can be tricky, since these are sometimes hidden from view.
Key concepts: APIs, API endpoints


Over the next two lectures, we’ll go over the main differences between the two approaches and cover the implications for any webscraping activity. I want to forewarn you that webscraping typically involves a fair bit of detective work. You will often have to adjust your steps according to the type of data you want, and the steps that worked on one website may not work on another. (Or even work on the same website a few months later). All this is to say that webscraping involves as much art as it does science.

The good news is that both server-side and client-side websites allow for webscraping.1 If you can see it in your browser, you can scrape it.


# Webscraping with rvest (server-side)

The primary R package that we’ll be using today is rvest (link), a simple webscraping library inspired by Python’s Beautiful Soup (link), but with extra tidyverse functionality. rvest is designed to work with webpages that are built server-side and thus requires knowledge of the relevant CSS selectors… Which means that now is probably a good time for us to cover what these are.

# Application 1: Wikipedia

Okay, let’s get to an application. Say that we want to scrape the Wikipedia page on the Men’s 100 metres world record progression.

First, open up this page in your browser. Take a look at its structure: What type of objects does it contain? How many tables does it have? Do these tables all share the same columns? What row- and columns-spans? Etc.

Once you’ve familiarised yourself with the structure, read the whole page into R using the rvest::read_html() function.

```{r}
# library(rvest) ## Already loaded

m100 = read_html("http://en.wikipedia.org/wiki/Men%27s_100_metres_world_record_progression") 
m100
```

As you can see, this is an XML document2 that contains everything needed to render the Wikipedia page. It’s kind of like viewing someone’s entire LaTeX document (preamble, syntax, etc.) when all we want are the data from some tables in their paper.


### Table 1: Pre-IAAF (1881–1912)

Let’s start by scraping the first table on the page, which documents the unofficial progression before the IAAF. The first thing we need to do is identify the table’s unique CSS selector. Here’s a GIF of me using SelectorGadget to do that.

As you can see, working through this iterative process yields “div+ .wikitable :nth-child(1)”. We can now use this unique CSS selector to isolate the pre-IAAF table content from the rest of the HTML document. The core rvest function that we’ll use to extract the table content is html_element(), before piping it on to html_table() to parse the HTML table into an R data frame.

```{r}
pre_iaaf = 
  m100 %>%
  html_element("div+ .wikitable :nth-child(1)") %>% ## select table element
  html_table()                                      ## convert to data frame

pre_iaaf
```

Great, it worked!

I’ll tidy things up a bit so that the data frame is easier to work with in R. First, I’ll use the janitor::clean_names() convenience function to remove spaces and capital letters from the column names. (Q: How else could we have done this?) Second, I’ll use the lubridate::mdy() function to convert the date string to a format that R actually understands.

```{r}
# library(janitor) ## Already loaded
# library(lubridate) ## Already loaded
 
pre_iaaf =
  pre_iaaf %>%
  clean_names() %>%         ## fix the column names
  mutate(date = mdy(date))  ## convert string to date format

pre_iaaf
```

Now that we have our cleaned pre-IAAF data frame, we could easily plot it. I’m going to hold off doing that until we’ve scraped the rest of the WR data. But first, an aside on browser inspection tools.

### Aside: Get CSS selectors via browser inspection tools

SelectorGadget is a great tool. But it isn’t available on all browsers and can involve more work than I’d like sometimes, with all that iterative clicking.3 I therefore wanted to mention an alternative (and very precise) approach to obtaining CSS selectors: Use the “inspect web element” feature of your browser.

Here’s a quick example using Google Chrome. First, I open up the inspect console (Ctrl+Shift+I, or right-click and choose “Inspect”). I then proceed to scroll over the source elements, until Chrome highlights the table of interest on the actual page. Once the table (or other element of interest) is highlighted, I can grab its CSS by right-clicking and selecting Copy -> Copy selector.

```{r}
abcd = 
  m100 %>%
  html_element("#mw-content-text > div.mw-parser-output > table:nth-child(8)") %>% ## select table element
  html_table()                                      ## convert to data frame

abcd
```


# Challenge

### Table 2: Pre-automatic timing (1912–1976)

```{r}
table1976 = 
  m100 %>%
  html_element("h3+ .wikitable :nth-child(1)") %>% ## select table element
  html_table()                                      ## convert to data frame

table1976
```
```{r}
table1976 =
  table1976 %>%
  clean_names() %>%         ## fix the column names
  mutate(date = mdy(date))  ## convert string to date format

table1976
```


### able 3: Modern Era (1977 onwards)

```{r}
table1977 = 
  m100 %>%
  html_element("#mw-content-text > div.mw-parser-output > table:nth-child(19)") %>% ## select table element
  html_table() %>%
  clean_names() %>%         ## fix the column names
  mutate(date = mdy(date))

table1977
```

### Combined eras plot

Let’s combine our three separate tables into a single data frame. I’ll use base R’s rbind() to bind by row and include only the variables that are common to all of the three data frames. For good measure, I’ll also add an extra column describing which era each record was recorded under.


```{r}
wr100 = 
  rbind(
    pre_iaaf %>% select(time, athlete, nationality, date) %>% mutate(era = "Pre-IAAF"),
    table1976 %>% select(time, athlete, nationality, date) %>% mutate(era = "Pre-automatic"),
    table1977 %>% select(time, athlete, nationality, date) %>% mutate(era = "Modern")
    )

wr100
```

All that hard works deserves a nice plot, don’t you think?

```{r}
wr100 %>%
  ggplot(aes(x=date, y=time, col=fct_reorder2(era, date, time))) + 
  geom_point(alpha = 0.7) +
  labs(
    title = "Men's 100m world record progression",
    x = "Date", y = "Time",
    caption = "Source: Wikipedia"
    ) +
  theme(legend.title = element_blank()) ## Switch off legend title
```

# Application 2: Craiglist

There are several features of the previous Wikipedia example that make it a good introductory application. Most notably, the HTML table format provides a regular structure that is easily coercible into a data frame (via html_table()). Oftentimes, however, the information that we want to scrape off the web doesn’t have this nice regular structure. For this next example, then, I’m going to walk you through a slightly more messy application: Scraping items from Craiglist.

The specific items that I’m going to scrape here are audio speakers for sale in my local city of Eugene. But you can adjust the relevant URL search parameters to your own preferences — cars in Chicago, concert tickets in Cleveland, etc. — and the same principles should carry through.

### Extract the text

We start as we always do by reading in the HTML.

```{r}
base_url = "https://eugene.craigslist.org/search/sss?query=speakers&sort=rel&srchType=T"

craiglist = read_html(base_url)
```

Next, we need to identify the CSS selectors in order to extract the relevant information from this page. Once again, this involves quite a lot of iterative clicking with SelectorGadget. I’ll spare you (and myself) another GIF. But here is a screenshot of the final result once I’ve isolated the elements of interest. As you can see, the relevant selector is “.result-hood , .result-date , .result-price , .hdrlnk”.


Now comes the first tweak relative to our previous example. Instead of using html_element(), we’ll use html_elements() (i.e. plural) to extract all of the matching elements.4 I’ll assign the resulting object as speakers, although I won’t try to coerce it to an R array just yet.

```{r}
speakers = 
  craiglist %>% 
  html_elements(".result-hood , .result-date , .result-price , .hdrlnk")
```


At this point, you may be tempted to pipe the speakers object to html_table() to create a data frame like we did with our Wikipedia example. Unfortunately, that won’t work here because we are dealing with free-form text rather than regular table structure.

```{r}
html_table(speakers)
```

Instead, we’ll parse it as simple text via html_text(). This will yield a vector of strings, which I’ll re-assign the same speakers object.

```{r}
speakers = html_text(speakers)  ## parse as text
head(speakers, 20)              ## show the first 20 entries
```


### Coercing to a data frame

We now have a bit of work on our hands to convert this vector of strings into a usable data frame. (Remember: Webscraping is as much art as it is science.) The general approach that we want to adopt is to look for some kind of “quasi-regular” structure that we can exploit.

For example, we can see from my screenshot above that each sale item tends to have five separate text fields. (Counter-clockwise from the top: price, listing date, description, price (again), and location.) Based on this, we might try to transform the vector into a (transposed) matrix with five columns and from there into a data frame.

```{r}
head(as.data.frame(t(matrix(speakers, nrow=5))))
```


Uh-oh. This approach isn’t going to work because not every sale item lists all five text fields. Quite a few are missing the location field, for instance.

Let’s try a different tack. The key point about the “quasi-regular” structure that we’re trying to exploit is that it needs to be present for every sale item. Looking again at the webpage/screenshot, can you think of something that meets that criteria?



How about the listing date text field? This is automatically populated by Craiglist and doesn’t rely on the vagaries of individual seller text. In the code chunk that follows, I’m going to exploit the fact that each item includes a listing date and use that as an anchor for creating individual row entries. Now, I have to admit that I’m going to use some tools that we haven’t covered yet — iteration and functions. We’ll get to these important topics later on in the course, so I don’t want you to stress about understanding the details of my code. The more important thing is the intuition, where I’m (a) identifying the listing date entries in the speakers vector, and (b) using these date entries as anchors to indicate breaks points between separate sale items. For good measure, I’m going to do this with data.table rather than the tidyverse. But that’s mostly a matter of personal taste; it would be easy to translate my code to the latter if that’s what you’d prefer.


```{r}
# library(data.table) ## Already loaded

dates = as.Date(speakers, format = '%b %d')  ## Try to coerce to date of form "Jan 01"
idates = which(!is.na(dates))                ## Get index of all the valid dates (i.e. non-NA)

## Iterate over our date index vector and then combine into a data.table. We'll 
## use the listing date to define the start of each new entry. Note, however, 
## that it usually comes second among the five possible text fields. (There is 
## normally a duplicate price field first.) So we have to adjust the way we 
## define the end of that entry; basically it's the next index position in the 
## sequence minus two.
speakers_dt =
  rbindlist(lapply(
    seq_along(idates),
    function(i) {
      start = idates[i]
      end = ifelse(i!=length(idates), idates[i+1]-2, tail(idates, 1))
      data.table(t(speakers[start:end]))
    }
    ), fill = TRUE) ## Use fill=TRUE arg so that rbindlist allocates 5 cols to each row

speakers_dt
```

Looks like it worked. Sweet.

This last bit of code is optional — and, again, I’m not going to explain myself much — but is just to tidy up the data table somewhat.

```{r}
names(speakers_dt) = c('date', 'description', 'price', 'location')

speakers_dt[, ':=' (date = as.Date(date, format = '%b %d'),
                    price = as.numeric(gsub('\\$|\\,', '', price)))]

## Because we only get the month and day, some entries from late last year may
## have inadvertently been coerced to a future date. Fix those cases.
speakers_dt[date>Sys.Date(), date := date - years(1)]

## Drop missing entries
speakers_dt = speakers_dt[!is.na(price)]

speakers_dt
```

As ever, let’s reward our efforts with a nice plot. I’ll add a few bells and whistles to this one, but this is most certainly optional.

```{r}
ggplot(speakers_dt, aes(date, price)) + 
  geom_point(aes(fill = price), show.legend = FALSE,
             shape = 21, colour = 'black', size = 2, stroke = 0.1) +
  scale_y_comma() + 
  scale_fill_viridis_c(option = 'magma', begin = 0.3, end = 0.9) +
  labs(title = 'Speakers for sale near Eugene, OR',
       caption = 'Source: Craigslist',
       x = 'Listing date', y = 'Price (USD)') +
  theme_modern_rc()
```



```{r}
### library(extrafont)
### extrafont::font_import()
```














