---
title: "webscraping_API"
output: pdf_document
---

```{r}
## Load and install the packages that we'll be using today
library(pacman)
p_load(tidyverse, httr, lubridate, hrbrthemes, janitor, jsonlite, fredr, 
               listviewer, usethis)
## My preferred ggplot2 plotting theme (optional)
theme_set(hrbrthemes::theme_ipsum())
```


# Client-side, APIs, and API endpoints

Recall that websites or applications that are built using a client-side framework typically involve something like the following steps:

You visit a URL that contains a template of static content (HTML tables, CSS, etc.). This template itself doesn’t contain any data.
However, in the process of opening the URL, your browser sends a request to the host server.
If your request if valid, then the server issues a response that fetches the necessary data for you and renders the page dynamically in your browser.
The page that you actually see in your browser is thus a mix of static content and dynamic information that is rendered by your browser (i.e. the “client”).
All of this requesting, responding and rendering takes places through the host application’s API (or Application Program Interface). Time for a student presentation to go over APIs in more depth…


### Student presentation: APIs

If you’re new to APIs or reading this after the fact, then I recommend this excellent resource from Zapier: An Introduction to APIs. It’s fairly in-depth, but you don’t need to work through the whole thing to get the gist. The summary version is that an API is really just a collection of rules and methods that allow different software applications to interact and share information. This includes not only web servers and browsers, but also software packages like the R libraries we’ve been using.1 Key concepts include:

Server: A powerful computer that runs an API.
Client: A program that exchanges data with a server through an API.
Protocol: The “etiquette” underlying how computers talk to each other (e.g. HTTP).
Methods: The “verbs” that clients use to talk with a server. The main one that we’ll be using is GET (i.e. ask a server to retrieve information), but other common methods are POST, PUT and DELETE.
Requests: What the client asks of the server (see Methods above).
Response: The server’s response. This includes a Status Code (e.g. “404” if not found, or “200” if successful), a Header (i.e. meta-information about the reponse), and a Body (i.e the actual content that we’re interested in).
Etc.

A bit more about API endpoints
A key point in all of this is that, in the case of web APIs, we can access information directly from the API database if we can specify the correct URL(s). These URLs are known as an API endpoints.

API endpoints are in many ways similar to the normal website URLs that we’re all used to visiting. For starters, you can navigate to them in your web browser. However, whereas normal websites display information in rich HTML content — pictures, cat videos, nice formatting, etc. — an API endpoint is much less visually appealing. Navigate your browser to an API endpoint and you’ll just see a load of seemingly unformatted text. In truth, what you’re really seeing is (probably) either JSON (JavaScript Object Notation) or XML (Extensible Markup Language).

You don’t need to worry too much about the syntax of JSON and XML. The important thing is that the object in your browser — that load of seemingly unformatted text — is actually very precisely structured and formatted. Moreover, it contains valuable information that we can easily read into R (or Python, Julia, etc.) We just need to know the right API endpoint for the data that we want.

Let’s practice doing this through a few example applications. I’ll start with the simplest case (no API key required, explicit API endpoint) and then work through some more complicated examples.


# Application 1: Trees of New York City

NYC Open Data is a pretty amazing initiative. Its mission is to “make the wealth of public data generated by various New York City agencies and other City organizations available for public use”. You can get data on everything from arrest data, to the location of wi-fi hotspots, to city job postings, to homeless population counts, to dog licenses, to a directory of toilets in public parks… The list goes on. I highly encourage you to explore in your own time, but we’re going to do something “earthy” for this first application: Download a sample of tree data from the 2015 NYC Street Tree Census.

I wanted to begin with an example from NYC Open Data, because you don’t need to set up an API key in advance.2 All you need to do is complete the following steps:


Open the web page in your browser (if you haven’t already done so).
You should immediately see the API tab. Click on it.
Copy the API endpoint that appears in the popup box.
Optional: Paste that endpoint into a new tab in your browser. You’ll see a bunch of JSON text, which you can render nicely using the JSONView browser extension that we installed earlier.


Now that we’ve located the API endpoint, let’s read the data into R. We’ll do so using the fromJSON() function from the excellent jsonlite package (link). This will automatically coerce the JSON array into a regular R data frame. However, I’ll go that little bit further and convert it into a tibble, since the output is nicer to work with.

```{r}
# library(jsonlite) ## Already loaded

nyc_trees = 
  fromJSON("https://data.cityofnewyork.us/resource/uvpi-gqnh.json") %>%
  as_tibble()
nyc_trees
```

Aside on limits: Note that the full census dataset contains nearly 700,000 individual trees. However, we only downloaded a tiny sample of that, since the API defaults to a limit of 1,000 rows. I don’t care to access the full dataset here, since I just want to illustrate some basic concepts. Nonetheless, if you were so inclined and read the docs, you’d see that you can override this default by adding ?$limit=LIMIT to the API endpoint. For example, to read in only the first five rows, you could use:

`## Not run
fromJSON("https://data.cityofnewyork.us/resource/nwxe-4ae8.json?$limit=5")
`

Getting back on track, let’s plot our tree data just to show it worked. One minor thing I want to point out is that jsonlite::fromJSON() automatically coerces everything into a character, so we’ll also need to convert some columns to numeric before we plot.

```{r}
nyc_trees %>% 
  select(longitude, latitude, stump_diam, spc_common, spc_latin, tree_id) %>% 
  mutate_at(vars(longitude:stump_diam), as.numeric) %>% 
  ggplot(aes(x=longitude, y=latitude, size=stump_diam)) + 
  geom_point(alpha=0.5) +
  scale_size_continuous(name = "Stump diameter") +
  labs(
    x = "Longitude", y = "Latitude",
    title = "Sample of New York City trees",
    caption = "Source: NYC Open Data"
    )
```

Not too bad. This would probably be more fun / impressive with an actual map of New York behind it. We’ll save that for the spatial lecture that’s coming up later in the course, though.

Again, I want to remind you that our first application didn’t require prior registration on the Open Data NYC website, or creation of an API key. This is atypical. Most API interfaces will only let you access and download data after you have registered an API key with them. This is especially true if you want to access an API linked to a federal agency or institution (Census, BEA, etc.). So let’s work through an application where an API key is required…


# Application 2: FRED data

For this second example application, I’m going to show you how to download the data underlying the above chart using the FRED API. In fact, I’ll go one better. First, I’ll show you how to download it yourself, so that you get an understanding of what’s happening underneath the hood. Then, I’ll direct you to a package that does all of the API work for you.

### Do it yourself


As with all APIs, a good place to start is the FRED API developer docs. If you read through these, you’d see that the endpoint path we’re interested in is series/observations. This endpoint “gets the observations or data values for an economic data series”. The endpoint documentation gives a more in-depth discussion, including the various parameters that it accepts.3 However, the parameters that we’ll be focused on here are simply:

file_type: “json” (Not required, but our preferred type of output.)
series_id: “GNPCA” (Required. The data series that we want.)
api_key: “YOUR_API_KEY” (Required. Go and fetch/copy your key now.)
Let’s combine these parameters with the endpoint path to view the data directly in our browser. Head over to https://api.stlouisfed.org/fred/series/observations?series_id=GNPCA&api_key=YOUR_API_KEY&file_type=json, replacing “YOUR_API_KEY” with your actual key. You should see something like the following: **93084ccc1a379c1378e37095b31bdee6**

At this point you’re probably tempted to read the JSON object directly into your R environment using the jsonlite::readJSON() function. And this will work. However, that’s not what we’re going to here. Rather, we’re going to go through the httr package (link). Why? Well, basically because httr comes with a variety of features that allow us to interact more flexibly and securely with web APIs.

Let’s start by defining some convenience variables such as the endpoint path and the parameters (which we’ll store in a list).


```{r}
endpoint = "series/observations"
params = list(
  api_key= "93084ccc1a379c1378e37095b31bdee6", ## Change to your own key
  file_type="json", 
  series_id="GNPCA"
  )
```

Next, we’ll use the httr::GET() function to request (i.e. download) the data. I’ll assign this to an object called fred.

```{r}
# library(httr) ## Already loaded above

fred = 
  httr::GET(
    url = "https://api.stlouisfed.org/", ## Base URL
    path = paste0("fred/", endpoint),    ## The API endpoint
    query = params                       ## Our parameter list
    )
```


To actually extract the content (i.e. data) from of this response, I’ll use the httr::content() function. Moreover, since we know that this content is a JSON array, we can again convert it into an R object using jsonlite::fromJSON().

```{r}
fred = 
  fred %>% 
  httr::content("text") %>% ## Extract the reponse content (i.e. text)
  jsonlite::fromJSON()      ## Convert from JSON to R object

## What type of object did we get?
typeof(fred)
```

It turns that the previous step has yielded a list object in R.4 so now we need to inspect this list to better understand its structure before extracting the information that we care about (and coerce it to a data frame.) I’d use the base View() function to do this in an interactive r session. but that won’t work as well for these lecture notes. Instead, I’ll use the listviewer::jsonedit() function to create an interactive widget that renders nicely in knitted r markdown documents.

```{r}
# View(fred) ## What I'd use in an interactive R session

## library(listviewer)        ## Already loaded
jsonedit(fred, mode = "view") ## Better for RMarkdown documents
```


Luckily, this particular list object isn’t too complicated. We can see that what we’re really interested in, is the fred$observations sub-element. I’ll use purrr::pluck() to extract this element (there are various other ways to do this) and then coerce it to a tibble.


```{r}
fred =
  fred %>% 
  purrr::pluck("observations") %>% ## Extract the "$observations" list element
  # .$observations %>% ## I could also have used this
  # magrittr::extract("observations") %>% ## Or this
  as_tibble() ## Just for nice formatting
fred
```

Okay! We’ve finally got our data and are nearly ready for some plotting. Recall that jsonlite::fromJSON() automatically converts everything to characters, so I’ll quickly change some variables to dates (using lubridate::ymd()) and numeric.

```{r}
# library(lubridate) ## Already loaded above

fred =
  fred %>%
  mutate(across(realtime_start:date, ymd)) %>%
  mutate(value = as.numeric(value)) 
```

plot

```{r}
fred %>%
  ggplot(aes(date, value)) +
  geom_line() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x="Date", y="2012 USD (Billions)",
    title="US Real Gross National Product", caption="Source: FRED"
    )
```

### Aside: Safely store and use API keys as environment variables

In the above example, I assumed that you would just replace the “YOUR_FRED_KEY” holder text with your actual API key. This is obviously not very secure or scalable, since it means that you can’t share your R script without giving away your key.5 Luckily, there’s an easy way to safely store and use sensitive information like API keys or passwords: Simply save them as an R environment variables. There are two, closely related approaches:

1. Set an environment variable for the current R session only.
2. Set an environment variable that persists across R sessions.
Let’s briefly review each in turn.

#### 1) Set an environment variable for the current R session only

Defining an environment variable for the current R session is very straightforward. Simply use the base Sys.setenv() function. For example:

```{r}
## Set new environment variable called MY_API_KEY. Current session only.
Sys.setenv(MY_API_KEY="93084ccc1a379c1378e37095b31bdee6") 
```

Once this is done, you can then safely assign your key to an object — including within an R Markdown document that you’re going to knit and share — using the Sys.getenv() function. For example:

```{r}
## Assign the environment variable to an R object
my_api_key = Sys.getenv("MY_API_KEY")
## Print it out just to show that it worked
my_api_key
```

Important: While this approach is very simple, note that in practice the Sys.setenv() part should only be run directly in your R console. Never include code chunks with sensitive Sys.setenv() calls in an R Markdown file or other shared documents.6 That would entirely defeat the purpose! Apart from the annoyance of having to manually set my API key each time I start a new R session, this is one reason that I prefer the next approach of persisting environment variables across sessions…


#### 2) Set an environment variable that persist across R sessions

The trick to setting an R environment variable that is available across sessions is to add it to a special file called ~/.Renviron. This is a text file that lives on your home directory — note the ~/ path — which R automatically reads upon startup. Because ~/.Renviron is just a text file, you can edit it with whatever is your preferred text editor. However, you may need to create it first if it doesn’t exist. A convenient way to do all of this from RStudio is with the usethis::edit_r_environ() function. You will need to run the next few lines interactively:

```{r}
## Open your .Renviron file. Here we can add API keys that persist across R sessions.
usethis::edit_r_environ() 
```

This will open up your ~/.Renviron file in a new RStudio window, which you can then modify as needed. As an example, let’s say that you want to add your FRED API key as an environment variable that persists across sessions. You can do this by simply adding a line like the below to your ~/.Renviron file and saving.7

`
FRED_API_KEY="abcdefghijklmnopqrstuvwxyz0123456789" ## Replace with your actual key
`

Once you have saved your changes, you’ll need to refresh so that this new environment variable is available in the current session. You could also restart R, but that’s overkill.

```{r}
## Optional: Refresh your .Renviron file.  
readRenviron("~/.Renviron") ## Only necessary if you are reading in a newly added R environment variable
```

Challenge: Once you’ve refreshed your ~/.Renviron file, try to re-download the FRED data from earlier. This time call your FRED API key directly as an environment variable in your parameter list using Sys.getenv() like this:

```{r}
params = list(
  api_key= Sys.getenv("FRED_API_KEY"), ## Get API directly and safely from the stored environment variable
  file_type="json", 
  series_id="GNPCA"
  )
```


# Application 3: World rugby rankings

Our final application will involve a more challenging case where the API endpoint is hidden from view. In particular, I’m going to show you how to access data on World Rugby rankings. Because — real talk — what’s more important than teaching Americans about rugby?

Disclaimer: World Rugby’s Terms & Conditions permits data downloading for own non-commerical use. It seems reasonable to me that these lecture notes fall under this use category.8 None of the methods presented below should be construed as an endorsement of data acquisition and use that violates these terms. Again: Just because you can scrape something, doesn’t mean you should.

Start by taking a look at the complicated structure of the website in a live session. Pay attention to the various tables and other interactive elements like calendars. Now take a minute or two for a quick challenge: Try to scrape the full country rankings using the rvest + CSS selectors approach that we practiced last time…

If you’re anything like me, you would have struggled to scrape the desired information using the rvest + CSS selectors approach. Even if you managed to extract some kind of information, you’re likely only getting a subset of what you wanted. (For example, just the column names, or the first ten rows before the “VIEW MORE RANKINGS” button). And we haven’t even considered trying to get information from a different date.9

### Locating the hidden API endpoint

Fortunately, there’s a better way: Access the full database of rankings through the API. First we have to find the endpoint, though. Here’s a step-by-step guide of how to that that. It’s fairly tedious, but pretty intuitive once you get the hang of it. You can just skip to the GIF below if you would rather see what I did instead of reading through all the steps.

Start by inspecting the page. (Ctr+Shift+I in Chrome. Ctrl+Shift+Q in Firefox.)
Head to the Network tab at the top of the inspect element panel.
Click on the XHR button.10
Refresh the page (Ctrl+R). This will allow us to see all the web traffic coming to and from the page in our inspect panel.
Our task now is to scroll these different traffic links and see which one contains the information that we’re after.
The top traffic link item references a URL called https://cmsapi.pulselive.com/rugby/rankings/mru?language=en&client=pulse. Hmmm. “API” you say? “Rankings” you say? Sounds promising…
Click on this item and open up the Preview tab.
In this case, we can see what looks to be the first row of the rankings table (“New Zealand”, etc.)
To make sure, you can grab that https://cmsapi.pulselive.com/rugby/rankings/mru?language=en&client=pulse, and paste it into our browser (using the JSONView plugin) from before.
Sweet. Looks like we’ve located our API endpoint. As promised, here’s GIF of me walking through these steps in my browser:

### Pulling the data into R

Let’s pull the data from the API endpoint into R. Again, I’ll be using jsonlite::readJSON() function.

```{r}
endpoint = "https://cmsapi.pulselive.com/rugby/rankings/mru?language=en&client=pulse"
rugby = fromJSON(endpoint)
str(rugby)
```


We have a nested list, where what looks to be the main element of interest, $entries, is itself a list.11 Let’s extract the $entries element and have a look at its structure. Again, I’d just use the base View() base in an interactive R session. Here I’m using listviewer::jsonedit() because it plays nicely with R Markdown documents.

```{r}
# str(rugby$entries) ## Base option
listviewer::jsonedit(rugby, mode = "view")
```


For completeness, let’s take a peak at the rugby$entries$team data frame to confirm that it has information that is useful to us.

```{r}
head(rugby$entries$team)
```

Okay, a clearer picture is starting to emerge. It looks like we can just bind the columns of the rugby$entries$team data frame directly to the other elements of the parent $team “data frame” (actually: “list”). Let’s do that using dplyr::bind_cols() and then clean things up a bit. I’m going to call the resulting data frame rankings.

```{r}
rankings =
  bind_cols(
    rugby$entries$team,
    rugby$entries %>% select(matches:previousPos)
    ) %>%
  clean_names() %>%
  select(-c(id, alt_id, annotations)) %>% ## These columns aren't adding much of interest
  select(pos, pts, everything()) %>% ## Reorder remaining columns
  as_tibble() ## "Enhanced" tidyverse version of a data frame
rankings
```

### BONUS: Get and plot the rankings history


NOTE: This bonus section involves some programming and loops. I know that we haven’t gotten to the programming section of the course, so don’t worry about the specifics of the next few code chunks. I’ll try to comment my code quite explicitly, but I mostly want you to focus on the big picture.

The above table looks great, except for the fact that it’s just a single snapshot of the most recent rankings. We are probably more interested in looking back at changes in the ratings over time.

But how to do this? Well, in the spirit of art-vs-science, let’s open up the Inspect window of the rankings page again and start exploring. What happens if we click on the calendar element, say, change the year to “2018” and month to “April”? (Do this yourself.)

This looks promising! Essentially, we get the same API endpoint that we saw previously, but now appended with a date, https://cmsapi.pulselive.com/rugby/rankings/mru?date=2018-05-01&client=pulse. If you were to continue along in this manner — clicking on the website calendar and looking for XHR traffic — you would soon realise that these date suffixes follow a predictable pattern: They are spaced out a week apart and always fall on a Monday. In other words, World Rugby updates its international rankings table weekly and publishes the results on Mondays.

We now have enough information to write a function that will loop over a set of dates and pull data from the relevant API endpoint. To start, we need a vector of valid dates to loop over. I’m going to use various functions from the lubridate package to help with this. Note that I’m only to extract a few data points — one observation a year for the last decade or so — since I only want to demonstrate the principle. No need to hammer the host server. (More on that below.)

```{r}
## We'll look at rankings around Jan 1st each year. I'll use 2004 as an
## arbitrary start year and then proceed until the present year.
start_date = ymd("2004-01-01")
end_date = floor_date(today(), unit="years")
dates = seq(start_date, end_date, by="years")
## Get the nearest Monday to Jan 1st to coincide with rankings release dates.
dates = floor_date(dates, "week", week_start = getOption("lubridate.week.start", 1))
dates
```

Next, I’ll write out a function that I’ll call rugby_scrape. This function will take a single argument: a date that it will use to construct a new API endpoint during each iteration. Beyond that, it will pretty do much exactly the same things that we did in our previous, manual data scrape. The only other difference is that it will wait three seconds after running (i.e. Sys.sleep(3)). I’m adding this final line to avoid hammering the server with instantaneous requests when we put everything into a loop.


```{r}
## First remove our existing variables. This is not really necessary, since R is smart enough
## to distinguish named objects in functions from named objects in our global environment.
## But I want to emphasise that we're creating new data here and avoid any confusion.
rm(rugby, rankings, endpoint)

## Now, create the function. I'll call it "rugby_scrape".
rugby_scrape = 
  function(x) {
    endpoint = paste0("https://cmsapi.pulselive.com/rugby/rankings/mru?date=", x, "&client=pulse")
    rugby = fromJSON(endpoint)
    rankings =
      bind_cols(
        rugby$entries$team,
        rugby$entries %>% select(matches:previousPos)
        ) %>%
      clean_names() %>%
      mutate(date = x) %>% ## New column to keep track of the date 
      select(-c(id, alt_id, annotations)) %>% ## These columns aren't adding much of interest
      select(date, pos, pts, everything()) %>% ## Reorder remaining columns
      as_tibble() ## "Enhanced" tidyverse version of a data frame
    Sys.sleep(3) ## Be nice!
    return(rankings)
  }
```

Finally, we can now iterate (i.e. loop) over our dates vector, by plugging the values sequentially into our rugby_scrape function. There are a variety of ways to iterate in R, but I’m going to use an lapply() call below.12 We’ll then bind everything into a single data frame using dplyr::bind_rows() and name the resulting object rankings_history.


```{r}
rankings_history =
  lapply(dates, rugby_scrape) %>% ## Run the iteration
  bind_rows() ## Bind the resulting list of data frames into a single data frame
rankings_history
```


Let’s review what we just did:

We created a vector of dates — creatively called dates — with observations evenly spaced (about) a year apart, falling on the Monday closest to Jan 1st for that year.
We then iterated (i.e. looped) over these dates using a function, rugby_scrape, which downloaded and cleaned data from the relevant API endpoint.
At the end of each iteration, we told R to wait a few seconds before executing the next step. Remember that R can execute these steps much, much quicker than we could ever type them manually. It probably doesn’t matter for this example, but you can easily “overwhelm” a host server by hammering it with a loop of automated requests. (Or, just as likely: They have safeguards against this type of behaviour and will start denying your requests as a suspected malicious attack.) As ever, the “be nice” motto holds sway when scraping web data.
Note that each run of our iteration will have produced a separate data frame, which lapply() by default appends into a list. We used dplyr::bind_rows() to bid these separate data frames into a single data frame.
Okay! Let’s plot the data and highlight a select few countries in the process.

```{r}
teams = c("NZL", "RSA", "ENG", "JPN")
team_cols = c("NZL"="black", "RSA"="#4DAF4A", "ENG"="#377EB8", "JPN" = "red")

rankings_history %>%
  ggplot(aes(x=date, y=pts, group=abbreviation)) +
  geom_line(col = "grey") + 
  geom_line(
    data = rankings_history %>% filter(abbreviation %in% teams), 
    aes(col=fct_reorder2(abbreviation, date, pts)),
    lwd = 1
    ) +
  scale_color_manual(values = team_cols) +
  labs(
    x = "Date", y = "Points", 
    title = "International rugby rankings", caption = "Source: World Rugby"
    ) +
  theme(legend.title = element_blank())
```










